Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2025-01-20 analysis/paper/article.qmd
JB Round 1 minor edits and responses to BM comments, 20 Jan 2025.

<!--Ben, please note that for now I am keeping track of my references. I have bolded them in text below; I think I will want to learn how to use Zotero and incorporate the citations using your arcane Lovecraftian ritual language (BibTex?), so that for our next paper I am capable of participating in the entire workflow with less supervision. BM: ok great! Yes, BibTeX, I believe you will love it. -->

@@

<!-- BM: I think we have a little repetition in the Intro and Background sections, particulary in the mentions of data from papers by Speakman, Kawa et al. and Mackie & Rockwell. To reduce duplication, I've removed most of that from the intro, and  kept/shifted it into the background so we have a logical progression of general to more specific observations about the job market -->

@@

In anthropology in the US the number of doctoral anthropology graduates has increased by about 70% over the past 30 years, but the number of new faculty positions has not increased proportionally [@speakman2018choosing]. New faculty positions have dwindled, in part due to the removal of the of the Age Discrimination in Employment Act (ADEA) exemption in 1994 which prohibited mandatory retirement ages in higher education (**Earle and Delpo Kulow 2015**). When combined with the institution of tenure in American academia, the ADEA exemption allowed [faculty to stay in their posts for as long as they liked] the median age for faculty in the U.S. now ranks among the highest for all professions (**Citation**). In tandem with the gradual de-investment in American higher education since the **XXX (Mirowski 2011**) and the aftershocks of the 2008 recession, the ADEA exemption has cultivated an environment where new lines are few and far between. Among biological anthropologists, @Passalacqua_2018 found a ratio of 0.81 PhDs to job academic advertisements per year, concluding that academic positions in biological anthropology are barely at sustainable levels. This echoes findings from other fields. In biomedical sciences, for example, there is one tenure-track position in the US for approximately every 6.3 PhD graduates [@Ghaffarzadegan_2015]. **Citation of Larson et al. 2014**.

PhD graduates who do not go directly into a tenure-track faculty position after graduation often go into a series of short-term appointments as part-time or limited contract instructors, also known as adjunct or contingent faculty. Rising numbers of contingent faculty has long been a concern in US higher education [@Trevithick2010]. Data from the American Association of University Professors' 2021–22 faculty survey indicate that more than 60% of faculty positions in US universities were held by non-tenure-track full time or part time contingent faculty members [@colby2022annual]. Contingent faculty positions are precarious because they provide low or no health and retirement benefits, little or nothing in the way of professional development, and lower salaries relative to tenure-track positions. PhD graduates who have a sequence of short-term academic positions are often disadvantaged financially, due to low compensation combined with expensive frequent relocation, socially due to isolation from family and community, and professionally due to being ineligible for many decision-making roles at the universities they work at [@platzerAcademicPrecarityAmerican2018].

Success in tenure-track job applications in archaeology is strongly determined by where applicants get their PhDs. Data from the 2014–2015 AnthroGuide publication of the American Anthropological Association shows that just ten out of over 100 US graduate programs produced over 30% of the graduates hired into tenure-track faculty positions [@speakman2018choosing]. Top-ranking programmes are placing fewer than one in three graduates in tenure-track jobs [@mackieMarketShareAccounting2023]. Similarly, a network analysis of 1,918 faculty holding tenured or tenure-track positions at PhD-granting anthropology programs in the US in 2015 (including 506 archaeologists) found that the just fifteen graduate programs produced 52.8 percent of tenured and tenure-track positions [@kawaSocialNetworkUS2019]. This network analysis showed that programs with large endowments and with faculty who hold prestigious awards and are widely cited by other scholars produce the majority of tenured and tenure-track faculty in anthropology. Hiring bias predicated on the prestige of specific institutions and programs is not unique to anthropology. Targeted studies of the disciplines of sociology, communication, and operations research, and industrial systems engineering show similar trends (**Barnett et al., 2010; Castillo et al., 2018; Feeley and Tutzauer 2022; Nevin 2019**). **Broader analyses comparing X, Y and Z, and analyses while a recent analysis of XXX showed Z (Clauset et al., 2015;** [@wapmanQuantifyingHierarchyDynamics2022a]). Within anthropology, however, such studies draw attention to the paradox of a discipline committed to fighting social inequalities continuing to reproduce its own systematic inequalities through hiring practices that favor an elite minority of graduates with prestigious institutional affiliations.

One way that some hiring committees are tackling these biases towards prestigious institutions is by providing detailed instructions to applicants on how to prepare their application materials. In theory, detailed applications will allow the hiring committee to focus on evaluating the accomplishments of candidates across common categories, rather than ranking based on prestige signals in a CV, such as a name of the applicant's graduate program. This push for equity has resulted in job ads that are often highly prescriptive in the types of documents that applicants should submit. For example, in addition to a cover letter and CV, job ads in many fields now require applicants to submit short statements detailing their previous and future contributions to teaching, research, and diversity within their discipline. In a comparison of job ads from 1999–2000 and 2019–2020 published in *Anthropology News* @HelloTristesTropes2021 noticed an increase in the number of materials requested from applicants. For example, twice as many 2019–2020 job ads requested writing samples compared to 1999–2000, and nearly four times as many requested statements of teaching philosophy. In their review of the 'worst job ads of 2021' @dennisWorstAnthroJob2022 reported job ads from Oberlin College and Grinnell College with particularly extreme requirements <!--Are you opposed to naming and shaming these institutions, which were Oberlin College and Grinnell College? I applied for one of these positions, and have no issue referring to them by name... BM: nope, I've added their names -->, which respectively requested nine and thirteen documents from applicants. This high burden on applicants disproportionately favors people with more time and financial resources to prepare the required materials.

A notable requirement that has emerged in recent years is a diversity statement, where the applicant describes their knowledge of, prior contributions to, and future goals for advancing diversity, equity, inclusion. This requirement has been much-debated largely due to an experiment in 2016–2022 at several University of California campuses that used diversity statements as the first cut for selecting candidates for tenure-track faculty positions [@soucek2021diversity]. Only candidates that scored highly on their diversity statements would have the rest of their application evaluated. This experiment generated intense and widespread public debate about the merits and risks of requiring and using diversity statements. These debates drew attention to the challenges of hiring faculty from underrepresented minorities, and resulted in a variety of approaches to evaluating diversity, even leading some universities to entirely omit requirements for diversity statements in job applications [@guidenAreDiversityStatements].

@@

Our primary data source is the Archaeology Academic Jobs Wiki<!--Check if Kelsky has anything too say about this. I also think there have been some articles about the pros and cons of the poli sci or Econ versions of this. BM: I did have a search of her blog where there are ~20 mentions, all pretty neutral, her most interesting comment is to notes that it originated in the year of the Great Recession -->. Originating in 2007, this is a set of freely accessible web pages that anyone can edit (anonymously or with a free user account) hosted by Fandom, a for-profit company. The Archaeology pages are part of the Academic Jobs Wiki, which coordinates similar collaboratively-edited resources for around 40 academic disciplines. The coordinators and contributors are nearly all anonymous or pseudonymous. Typically, contributors copy and paste all or some of the text of job ads from other sources, such as the *Chronicle of Higher Education*, *Higher Ed Jobs*, and university websites, into the wiki, collecting <!--consolidating? centralizing? BM: I prefer the shortest and simplest word that will do the job --> ads originally posted in numerous different locations. Other contributors then edit the web page to add comments below an ad to share relevant information based on their experience in applying for that position. These edits result in annotations such as a tally of how many people have applied, the dates of events such as requests for more materials, interviews, offer made, rejection notices, etc. Contributors also edit the page to ask and answer questions about the positions and the application process. These comments make the Academic Jobs Wiki a unique resource for timely and specific information for job-seekers about positions they are interested in, and one of the most important internet resources for the academic job market. Because of its reputation for aggregating ads from diverse sources and rapidly-updated information that is not available elsewhere, the Academic Jobs Wiki has a large community of users that keep it updated and accurate has become an authoritative data source for studies of hiring trends in academia [e.g. @Passalacqua_2018; @musial2018five] and a widely recommended resource [e.g. @lightfoot2021preparing].

For each tenure-track job advertised on the Archaeology Academic Jobs Wiki during 2013-2023, we read the text and recorded the name of the hiring institution, the title of the position, and exact words and phrases from the ad about the topical, geographic, and methods foci on the position into a Google form. The topical focus is what we understood as the primary intellectual focus of the position—examples of topical foci included environmental archaeology, public archaeology, and North American archaeology. The geographic focus is the region of the world that the ideal candidate has scholarly expertise on, for example, Southwest US, Mediterranean, or Asia and India. The methods focus is the data-generating sub-field of archaeology that is mentioned in the ad. Examples of methods used in this study include archaeobotany, lithic analysis, or zooarchaeology. <!--It might be helpful to make this into a summary table with examples of all foci generated by the study, and then one showing the simplified 10–15 topics. I understand this is in the supplementary tables already. BM: I've added in the 'for example' you requested, but feels very repetitive to give examples in this paragraph, and then the full set of topics etc. in the next paragraph, and add then add a table as well. It's like saying "we found 5000 lithics, one random one was 45 mm long, another random piece was 12 mm wide, etc." readers don't expect that in the manuscript  --> We also recorded the type and number of documents requested in each ad (e.g. cover letter, CV, statements on research, teaching, diversity <!-- Were these all relatively easy to recognise? I’ve applied to places like Penn State, for example, that do a stupid combined research/teaching/diversity statement, just to make applicants do extra work, seemingly. BM: we treated these kinds of combo statements as separate requests, I don't know how many there were, I'm ok to omit that detail -->, syllabi, course descriptions, writing samples, transcripts) and how many names/letters of recommenders were requested in the ad, if any.

@@

We collected data from `r total_number_of_ads_in_our_sample` ads for tenure-track jobs in archaeology posted during 2013-2023. We focus our analysis here on the `r total_number_of_us_ads_in_our_sample` ads for positions at US universities <!--Ben, if we should provide the rationale for this, I don't remember what it was BM: I also don't recall, I don't see any justification for the US focus in Speakman et al or Kawa et al. and I can't think of a one that is either unverifiable or doesn't require heaps of additional work to validate. Leave it to the peer reviewers? -->. @fig-show-basic-plots shows the count of ads for each year. Assistant Professor jobs are consistently the most common title and rank of positions advertised, while open rank or full professor are the least frequent. The ratio of tenure-track to non-tenure-track positions is generally well above one; in other words, this data set is dominated by tenure-track positions . Only academic year 2013-2014 had more non-tenure-track positions than tenure-track positions <!--I know that this is a pedantic complaint, but we should be consistent about whether or not we hyphenate tenure-track BM: I've edited every instance so they have hyphens. How do you feel about the double hyphenated 'non-tenure-track'? -->, which was followed by an upward trend peaking at 2018-2019 and then declining again into the present.

@@

Panel B of @fig-show-map-of-hiring-institution-and-cc shows the geographic distribution of the hiring institutions. California posted almost twice as many job ads as the next most active states. After California, the states that posted the most ads during 2013-2023 include New York, Texas, and Pennsylvania, and Florida. These top five states also correspond to the top five most populous US states, suggesting that rates of hiring are approximately proportional to population density. These top five states for job ads are also the top five states with highest number of degrees awarded in Anthropology (U.S. Department of Education nd). Similarly, the lowest counts of job ads were observed in states with the lowest populations: North Dakota, South Dakota, Alaska, and Nebraska. No institutions in Montana posted a job ad during this period. The implication here is that job-seekers who are able to relocate to populous areas will have more employment options. <!--Yes, nicely put BM: thanks! -->

@@

We recorded all geographic regions mentioned in the text where the successful applicant should have expertise and be research active. Our analysis focuses on those locations mentioned in 20 or more ads. Overall, American locations dominate. Panel A of @fig-show-geo-trends shows that a single region of the US, the Southwest, occurs in more job ads than every other part of the world except for the Mediterranean. <!-- Someone in another field recently asked me what archaeologists wear to conferences, and I said “bolo ties”. This is why. BM: haha, yes, exactly! -->The Southwest includes Arizona and New Mexico, with portions of California, Colorado, Nevada, Oklahoma, Texas, and Utah. It is archaeologically significant as the home of the Ancestral Pueblo, Hohokam, and Mogollon peoples who practiced irrigation agriculture and lived in relatively large settlements, compared to other regions of the US. The area was later occupied by the Navajo, Ute, Southern Paiute, Hopi and Zuni, also in relatively large numbers [@griffin2000native] <!-- BM: added requested citation: JB: Summary citation(s), preferably including an Indigenous archaeologist from the region -->. The Mediterranean is prominent because it is the location that is often mentioned in job ads focused on classical archaeology (i.e. archaeology of Bronze and Iron Age Italy and Greece).

@@

Landscape archaeology, encompassing GIS and remote sensing, has remained prominent compared to other methods [@fig-show-metho-trends]. <!--Might be worth exploring whether this is a skillset more frequently sought after in CRM and/or other industries—is this a transferable skill thing? BM: yes, good point, added something on that --> The popularity of this suite of methods may reflect its importance as a skill in demand by employers in the Cultural Research Management sector. @morganReadyNotArchaeological2023 found that almost one-third of 599 jobs ads posted by CRM employers sought candidates with experience in GIS methods. Methods focused on a specific element of the archaeological record, such as Lithic analysis, Zooarchaeology and Ceramics are among the least frequently mentioned in job ads. Instead, the more popular methods are those that are relevant to multiple elements of the archaeological record (e.g. Archaeobotany encompasses macroscopic and microscopic plant remains; Bioarchaeology may include skeletal analysis, isotopes, proteins, etc.).

Landscape archaeology, although dominant, has fluctuated over the years and has been on a downtrend since 2018-2019. Computational and digital archaeology is the second most represented method, showing an overall increasing trend, particularly since 2020-2021. <!--Congratulations BM: haha, would you like to borrow my crystal ball?--> Archaeobotany shows a strong cyclical trend, rising, falling, and then rising again over our study period. Archaeometry and Geoarchaeology have maintained a relatively low but steady presence in job ads, peaking in 2017-2018 and 2018-2019 and declining thereafter. Lithic analysis and Zooarchaeology are also mentioned relatively infrequently in job ads and show an inverse correlation with each other after 2018-2019.

@@

The most frequently mentioned topic in the job ads is Environmental archaeology [@fig-show-topi-trends]. This category encompasses such phrases as human-environmental dynamics, interaction between humans and their environments, environmental change, climate change, historical ecology, ecological knowledge,human ecology, and ecological systems. Public archaeology is the second most frequent topic overall, which included phrases such as cultural resource management, cultural heritage, heritage studies, museum studies, human rights, community engaged, historic preservation, social justice, community-based, repatriation, and community-engaged archaeology. The least common topics in our sample are Pleistocene archaeology, e.g. human origins, hunter-gatherer archaeology, and digital archaeology. <!-- BM: See, here is a nice unpacking of some of the key categories -->

In the years 2019-2020 and 2020-2021 there are striking changes in the popularity of topics in job ads. Indigenous and historical archaeology (which includes archaeology of the African diaspora and enslaved people <!--African diaspora archaeology? We should be careful to use the appropriate language here BM: yes, ok, added that, but many jobs ads literally mention slavery and enslaved people, so I want to make it clear that slavery is a notable topic in American archaeology, not something to sanitize and ignore under the gloss of historical archaeology generally being about white people, a personal obsession, I guess. Did you know there were African ex-slaves on the first fleet to Australia? -->) became the most popular topic at this time, rising from being one of the least popular topics from 2012-2017. Conversely, archaeological science, which was popular during 2012-2017, was rarely mentioned in job ads during 2019-2021. Ancient Europe and the Mediterranean, which was an infrequently mentioned topic for the entire study period, virtually disappeared from job ads during 2019-2021. Mentions of complex societies in ads decrease at a steady rate, while mentions of public archaeology consistently increase over the study period. The topic of Environmental Archaeology remains <!--We should be careful of the word consistent as we are relying on it a bit heavily throughout the text. Here, could we replace with "popular" or "stable"? BM: yes, good point, seems to work if I just delete it -->high over time.

@@

Over our ten year study period there have been substantial changes in the instructions to applicants in terms of the type and number of documents that are requested by the search committees [@fig-requirements-over-time]. Requests for cover letter and CV decline in more recent years, perhaps reflecting their status as a generic and expected component to submit without needing to be specifically requested <!--So curious as to what these jobs DO request. BM: I'm not confident this is a real or important trend, but I did look into the schools that don't ask for these, there are mostly lesser-known places like Texas Tech University and East Tennessee State University. I think there might be a few other factors here, for example the job ads on the wiki might have that generic text edited out because it's expected by everyone. Difficult to test this because the links the interfolio, etc original source text for most ads are broken now, I did check a dozen of them just now. In any case, I don't think any further work on it would be of much relevance to our main arguments -->. The requirement for a diversity statement is rare until 2019-2020, peaks around 2020-2021, then decreases towards the present. Requests for names of recommenders (either zero or three names, rarely only two names) reaches a maximum during 2019-2020, then also decline for the remainder of the study period. The requirement for a research statement and teaching statement increases after 2015-2016, and becomes more frequent in job ads in more recent years. Requests for course descriptions, syllabus samples, teaching evaluations, transcripts and writing samples are consistently low over time (not shown here).

@@

Our data on the requirements for applicants support prior findings that the complexity of applications—and concomitantly, the labor required to apply for a tenure-track job—has gradually increased over time. This trend is especially pronounced for Assistant Professor positions, which make more demands on applicants than Associate and Full Professor positions. <!--Eat the old-->. Consistent with results from other studies \[\@HelloTristesTropes2021\], our study documented a growing demand for research and teaching statements. Demand for diversity statements shows a unique trajectory, peaking in 2020-2021 and declining into the present. This may relate to the intersecting concerns emerging during the COVID-19 pandemic, which may have reached a peak in 2020-2021 and then declined over time, resulting in dropping requests for diversity statements, as the most urgent period of the pandemic moves into the past. <!-- I don't understand the second half of this sentence BM: edited in an attempt to clarify -->. Another factor here may be the debates surrounding the experiment with diversity statements in hiring at some UC campuses during 2016-2022. Universities that previously required a diversity statement dropped that requirement after 2022 [@guidenAreDiversityStatements]. In our data we observed a reversal of the diversity statement requirement at `r n_schools_that_dropped_ds_req` universities. Those schools posted job ads prior 2022 that did require a diversity statement and also posted an ad in 2022 that did not require one.

One bright spot for applicants is the decline in recent years for requests for names of recommenders in the initial application. This may be a response to recent criticisms of the burden on the applicant of preparing numerous complex job applications [e.g. @dennisWorstAnthroJob2022]. In recognition of this burden, not only on applicants but also on colleagues writing letters of recommendation over and over to support applicants, many hiring committees now follow the recommendations of @dennisWorstAnthroJob2022 in only requesting names and letters of recommendation at later stages of the hiring process, if at all. Showing sensitivity to this burden, in 2020 the American Anthropological Association issued guidance to academic departments that letters of recommendation should not be requested in the initial application, but only from short-listed candidates [@mallinLetsMakeAcademic2020] <!-- A line here about AAA recommendations to not ask for letters up front. BM: I looked for this statement and could only find broken links, so not sure if they are still into it. Added something anyway. -->

@@

We thank NO ONE. <!-- haha! so generous of you! -->

>>>>

@@

<!-- BM: I think we have a little repetition in the Intro and Background sections, particulary in the mentions of data from papers by Speakman, Kawa et al. and Mackie & Rockwell. To reduce duplication, I've removed most of that from the intro, and  kept/shifted it into the background so we have a logical progression of general to more specific observations about the job market. JB: Duly noted, thanks for doing the legwork on that. -->

@@

In anthropology in the US the number of doctoral anthropology graduates has increased by about 70% over the past 30 years, but the number of new faculty positions has not increased proportionally [@speakman2018choosing]. New faculty positions have dwindled, in part due to the removal of the of the Age Discrimination in Employment Act (ADEA) exemption in 1994 which prohibited mandatory retirement ages in higher education (**Earle and Delpo Kulow 2015**). When combined with the institution of tenure in American academia, the ADEA exemption allowed faculty to stay in their posts for as long as they liked. The median age for faculty in the U.S. now ranks among the highest for all professions (**Citation**). In tandem with the gradual de-investment in American higher education since 1980s **(Mirowski 2011**) and the aftershocks of the 2008 recession, the ADEA exemption has cultivated an environment where new lines are few and far between. Among biological anthropologists, @Passalacqua_2018 found a ratio of 0.81 PhDs to job academic advertisements per year, concluding that academic positions in biological anthropology are barely at sustainable levels. This echoes findings from other fields. In biomedical sciences, for example, there is one tenure-track position in the US for approximately every 6.3 PhD graduates \[@Ghaffarzadegan_2015\]. In engineering, Larson et al., have calculated that providing jobs for even 50% of PhD graduates would require the field to expand at an "improbable" rate of 14% per year (**2014: 747**). As they emphasize, "... the system in many places is saturated, far beyond capacity to absorb new PhDs in academia at the rates they are being produced" (**2014: 749**).

PhD graduates who do not go directly into a tenure-track faculty position after graduation often go into a series of short-term appointments as part-time or limited contract instructors, also known as adjunct or contingent faculty. Rising numbers of contingent faculty has long been a concern in US higher education [@Trevithick2010]. Data from the American Association of University Professors' 2021–22 faculty survey indicate that more than 60% of faculty positions in US universities were held by non-tenure-track full time or part time contingent faculty members [@colby2022annual]. Contingent faculty positions are precarious because they provide low or no health and retirement benefits, limited opportunities for professional development, and lower salaries relative to tenure-track positions. PhD graduates who have a sequence of short-term academic positions are often disadvantaged financially, due to low compensation combined with expensive frequent relocation, socially due to isolation from family and community, and professionally due to being ineligible for many decision-making roles at the universities at which they work [@platzerAcademicPrecarityAmerican2018]. <!--Well, this is depressingly familiar. -->

Success in tenure-track job applications in archaeology is strongly determined by where applicants get their PhDs. Data from the 2014–2015 AnthroGuide publication of the American Anthropological Association shows that just ten out of over 100 US graduate programs produced over 30% of the graduates hired into tenure-track faculty positions [@speakman2018choosing]. Top-ranking programs are placing fewer than one in three graduates in tenure-track jobs [@mackieMarketShareAccounting2023]. Similarly, a network analysis of 1,918 faculty holding tenured or tenure-track positions at PhD-granting anthropology programs in the US in 2015 (including 506 archaeologists) found that the just fifteen graduate programs produced 52.8 percent of tenured and tenure-track positions [@kawaSocialNetworkUS2019]. This network analysis showed that programs with large endowments, with faculty who hold prestigious awards and are widely cited by other scholars, produce the majority of tenured and tenure-track faculty in anthropology. Hiring bias predicated on the prestige of specific institutions and programs is not unique to anthropology. Targeted studies of the disciplines of sociology, communication, and operations research, and industrial systems engineering show similar trends (**Barnett et al., 2010; Castillo et al., 2018; Feeley and Tutzauer 2022; Nevin 2019**). **Broader analyses comparing X, Y and Z, and analyses while a recent analysis of XXX showed Z (Clauset et al., 2015;** [@wapmanQuantifyingHierarchyDynamics2022a]). Within anthropology, however, such studies draw attention to the paradox of a discipline committed to fighting social inequalities continuing to reproduce its own systematic inequalities through hiring practices that favor an elite minority of graduates with prestigious institutional affiliations.

One way that some hiring committees are tackling these prestige biases is by providing detailed instructions to applicants on how to prepare their application materials. In theory, detailed applications will allow the hiring committee to focus on evaluating the accomplishments of candidates across common categories, rather than ranking based on prestige signals in a CV, such as a name of the applicant's graduate program. This push for equity has resulted in job ads that are often highly prescriptive in the types of documents that applicants should submit. For example, in addition to a cover letter and CV, job ads in many fields now require applicants to submit short statements detailing their previous and future contributions to teaching, research, and diversity within their discipline. In a comparison of job ads from 1999–2000 and 2019–2020 published in *Anthropology News* @HelloTristesTropes2021 noticed an increase in the number of materials requested from applicants. For example, twice as many 2019–2020 job ads requested writing samples compared to 1999–2000, and nearly four times as many requested statements of teaching philosophy. In their review of the 'worst job ads of 2021' @dennisWorstAnthroJob2022 reported job ads from Oberlin College and Grinnell College with particularly extreme requirements, which respectively requested nine and thirteen documents from applicants. This high burden on applicants disproportionately favors people with more time and financial resources to prepare the required materials.

A notable requirement that has emerged in recent years is a diversity statement, where the applicant describes their knowledge of, prior contributions to, and future goals for advancing diversity, equity, and inclusion. This requirement has been much-debated largely due to an experiment in 2016–2022 at several University of California campuses that used diversity statements as the first cut for selecting candidates for tenure-track faculty positions [@soucek2021diversity]. Only candidates that scored highly on their diversity statements would have the rest of their application evaluated. This experiment generated intense and widespread public debate about the merits and risks of requiring and using diversity statements. These debates drew attention to the challenges of hiring faculty from underrepresented minorities, and resulted in a variety of approaches to evaluating diversity, even leading some universities to entirely omit requirements for diversity statements in job applications [@guidenAreDiversityStatements]. <!--The Harvard PhD student staying with me right now told me that Harvard just dropped their diversity statement requirement, if we want a specific example. Not sure it's necessary, though. -->

@@

Our primary data source is the Archaeology Academic Jobs Wiki<!--Check if Kelsky has anything too say about this. I also think there have been some articles about the pros and cons of the poli sci or Econ versions of this. BM: I did have a search of her blog where there are ~20 mentions, all pretty neutral, her most interesting comment is to notes that it originated in the year of the Great Recession -->. Originating in 2007, this is a set of freely accessible web pages that anyone can edit (anonymously or with a free user account) hosted by Fandom, a for-profit company. The Archaeology pages are part of the Academic Jobs Wiki, which coordinates similar collaboratively-edited resources for around 40 academic disciplines. The coordinators and contributors are nearly all anonymous or pseudonymous. Typically, contributors copy and paste all or some of the text of job ads from other sources, such as the *Chronicle of Higher Education*, *Higher Ed Jobs*, and university websites, into the wiki, collecting <!--consolidating? centralizing? BM: I prefer the shortest and simplest word that will do the job. JB: How very Strunk & White of you. We'll leave it as collective.--> ads originally posted in numerous different locations. Other contributors then edit the web page to add comments below an ad to share relevant information based on their experience in applying for that position. These edits result in annotations such as a tally of how many people have applied, the dates of events such as requests for more materials, interviews, offer made, rejection notices, etc. Contributors also edit the page to ask and answer questions about the positions and the application process. These comments make the Academic Jobs Wiki a unique resource for timely and specific information for job-seekers about positions they are interested in, and one of the most important internet resources for the academic job market. Because of its reputation for aggregating ads from diverse sources and rapidly-updated information that is not available elsewhere, the Academic Jobs Wiki has a large community of users that keep it updated and accurate has become an authoritative data source for studies of hiring trends in academia [e.g. @Passalacqua_2018; @musial2018five] and a widely recommended resource [e.g. @lightfoot2021preparing].

For each tenure-track job advertised on the Archaeology Academic Jobs Wiki during 2013-2023, we read the text and recorded the name of the hiring institution, the title of the position, and exact words and phrases from the ad about the topical, geographic, and methods foci on the position into a Google form. The topical focus is what we understood as the primary intellectual focus of the position—examples of topical foci included environmental archaeology, public archaeology, and North American archaeology. The geographic focus is the region of the world that the ideal candidate has scholarly expertise on, for example, Southwest US, Mediterranean, or Asia and India. The methods focus is the data-generating sub-field of archaeology that is mentioned in the ad. Examples of methods used in this study include archaeobotany, lithic analysis, or zooarchaeology. <!--It might be helpful to make this into a summary table with examples of all foci generated by the study, and then one showing the simplified 10–15 topics. I understand this is in the supplementary tables already. BM: I've added in the 'for example' you requested, but feels very repetitive to give examples in this paragraph, and then the full set of topics etc. in the next paragraph, and add then add a table as well. It's like saying "we found 5000 lithics, one random one was 45 mm long, another random piece was 12 mm wide, etc." readers don't expect that in the manuscript  --> We also recorded the type and number of documents requested in each ad (e.g. cover letter, CV, statements on research, teaching, diversity , syllabi, course descriptions, writing samples, transcripts) and how many names/letters of recommenders were requested in the ad, if any.

@@

We collected data from `r total_number_of_ads_in_our_sample` ads for tenure-track jobs in archaeology posted during 2013-2023. We focus our analysis here on the `r total_number_of_us_ads_in_our_sample` ads for positions at US universities . @fig-show-basic-plots shows the count of ads for each year. Assistant Professor jobs are consistently the most common title and rank of positions advertised, while open rank or full professor are the least frequent. The ratio of tenure-track to non-tenure-track positions is generally well above one; in other words, this data set is dominated by tenure-track positions . Only academic year 2013-2014 had more non-tenure-track positions than tenure-track positions , which was followed by an upward trend peaking at 2018-2019 and then declining again into the present.

@@

Panel B of @fig-show-map-of-hiring-institution-and-cc shows the geographic distribution of the hiring institutions. California posted almost twice as many job ads as the next most active states. After California, the states that posted the most ads during 2013-2023 include New York, Texas, and Pennsylvania, and Florida. These top five states also correspond to the top five most populous US states, suggesting that rates of hiring are approximately proportional to population density. These top five states for job ads are also the top five states with highest number of degrees awarded in Anthropology (U.S. Department of Education nd). Similarly, the lowest counts of job ads were observed in states with the lowest populations: North Dakota, South Dakota, Alaska, and Nebraska. No institutions in Montana posted a job ad during this period. The implication here is that job-seekers who are able to relocate to populous areas will have more employment options.

@@

We recorded all geographic regions mentioned in the text where the successful applicant should have expertise and be research active. Our analysis focuses on those locations mentioned in 20 or more ads. Overall, American locations dominate. Panel A of @fig-show-geo-trends shows that a single region of the US, the Southwest, occurs in more job ads than every other part of the world except for the Mediterranean. The Southwest includes Arizona and New Mexico, with portions of California, Colorado, Nevada, Oklahoma, Texas, and Utah. It is archaeologically significant as the home of the Ancestral Pueblo, Hohokam, and Mogollon peoples who practiced irrigation agriculture and lived in relatively large settlements, compared to other regions of the US. The area was later occupied by the Navajo, Ute, Southern Paiute, Hopi and Zuni, also in relatively large numbers [@griffin2000native] <!-- BM: added requested citation: JB: Summary citation(s), preferably including an Indigenous archaeologist from the region. JB: Thanks Ben, though this was a note for myself, not you :/-->. The Mediterranean is prominent because it is the location that is often mentioned in job ads focused on classical archaeology (i.e. archaeology of Bronze and Iron Age Italy and Greece).

@@

Landscape archaeology, encompassing GIS and remote sensing, has remained prominent compared to other methods [@fig-show-metho-trends]. The popularity of this suite of methods may reflect its importance as a skill in demand by employers in the Cultural Research Management sector. @morganReadyNotArchaeological2023 found that almost one-third of 599 jobs ads posted by CRM employers sought candidates with experience in GIS methods. Methods focused on a specific element of the archaeological record, such as Lithic analysis, Zooarchaeology and Ceramics are among the least frequently mentioned in job ads. Instead, the more popular methods are those that are relevant to multiple elements of the archaeological record (e.g. Archaeobotany encompasses macroscopic and microscopic plant remains; Bioarchaeology may include skeletal analysis, isotopes, proteins, etc.).

Landscape archaeology, although dominant, has fluctuated over the years and has been on a downtrend since 2018-2019. Computational and digital archaeology is the second most represented method, showing an overall increasing trend, particularly since 2020-2021. Archaeobotany shows a strong cyclical trend, rising, falling, and then rising again over our study period. Archaeometry and Geoarchaeology have maintained a relatively low but steady presence in job ads, peaking in 2017-2018 and 2018-2019 and declining thereafter. Lithic analysis and Zooarchaeology are also mentioned relatively infrequently in job ads and show an inverse correlation with each other after 2018-2019.

@@

The most frequently mentioned topic in the job ads is Environmental archaeology [@fig-show-topi-trends]. This category encompasses such phrases as human-environmental dynamics, interaction between humans and their environments, environmental change, climate change, historical ecology, ecological knowledge,human ecology, and ecological systems. Public archaeology is the second most frequent topic overall, which included phrases such as cultural resource management, cultural heritage, heritage studies, museum studies, human rights, community engaged, historic preservation, social justice, community-based, repatriation, and community-engaged archaeology. The least common topics in our sample are Pleistocene archaeology, e.g. human origins, hunter-gatherer archaeology, and digital archaeology.

In the years 2019-2020 and 2020-2021 there are striking changes in the popularity of topics in job ads. Indigenous and historical archaeology (which includes archaeology of the African diaspora and enslaved people <!--African diaspora archaeology? We should be careful to use the appropriate language here BM: yes, ok, added that, but many jobs ads literally mention slavery and enslaved people, so I want to make it clear that slavery is a notable topic in American archaeology, not something to sanitize and ignore under the gloss of historical archaeology generally being about white people, a personal obsession, I guess. Did you know there were African ex-slaves on the first fleet to Australia? -->) became the most popular topic at this time, rising from being one of the least popular topics from 2012-2017. Conversely, archaeological science, which was popular during 2012-2017, was rarely mentioned in job ads during 2019-2021. Ancient Europe and the Mediterranean, which was an infrequently mentioned topic for the entire study period, virtually disappeared from job ads during 2019-2021. Mentions of complex societies in ads decrease at a steady rate, while mentions of public archaeology consistently increase over the study period. The topic of Environmental Archaeology remains high over time.

@@

Over our ten year study period there have been substantial changes in the instructions to applicants in terms of the type and number of documents that are requested by the search committees [@fig-requirements-over-time]. Requests for cover letter and CV decline in more recent years, perhaps reflecting their status as a generic and expected component to submit without needing to be specifically requested. The requirement for a diversity statement is rare until 2019-2020, peaks around 2020-2021, then decreases towards the present. Requests for names of recommenders (either zero or three names, rarely only two names) reaches a maximum during 2019-2020, then also decline for the remainder of the study period. The requirement for a research statement and teaching statement increases after 2015-2016, and becomes more frequent in job ads in more recent years. Requests for course descriptions, syllabus samples, teaching evaluations, transcripts and writing samples are consistently low over time (not shown here).

@@

Our data on the requirements for applicants support prior findings that the complexity of applications—and concomitantly, the labor required to apply for a tenure-track job—has gradually increased over time. This trend is especially pronounced for Assistant Professor positions, which make more demands on applicants than Associate and Full Professor positions. Consistent with results from other studies \[\@HelloTristesTropes2021\], our study documented a growing demand for research and teaching statements. Demand for diversity statements shows a unique trajectory, peaking in 2020-2021 and declining into the present. This may relate to the intersecting concerns emerging during the COVID-19 pandemic, which may have reached a peak in 2020-2021 and then declined over time, resulting in dropping requests for diversity statements, as the most urgent period of the pandemic moves into the past. Another factor here may be the debates surrounding the experiment with diversity statements in hiring at some UC campuses during 2016-2022. Universities that previously required a diversity statement dropped that requirement after 2022 [@guidenAreDiversityStatements]. In our data we observed a reversal of the diversity statement requirement at `r n_schools_that_dropped_ds_req` universities. Those schools posted job ads prior 2022 that did require a diversity statement and also posted an ad in 2022 that did not require one.

One bright spot for applicants is the decline in recent years for requests for names of recommenders in the initial application. This may be a response to recent criticisms of the burden on the applicant of preparing numerous complex job applications [e.g. @dennisWorstAnthroJob2022]. In recognition of this burden, not only on applicants but also on colleagues writing letters of recommendation over and over to support applicants, many hiring committees now follow the recommendations of @dennisWorstAnthroJob2022 in only requesting names and letters of recommendation at later stages of the hiring process, if at all. Showing sensitivity to this burden, in 2020 the American Anthropological Association issued guidance to academic departments that letters of recommendation should not be requested in the initial application, but only from short-listed candidates [@mallinLetsMakeAcademic2020] <!-- A line here about AAA recommendations to not ask for letters up front. BM: I looked for this statement and could only find broken links, so not sure if they are still into it. Added something anyway. JB: I am finding it here under Job Board Policies: https://employers.americananthro.org/static-page/10285/job-board-policies/?-->

@@

We thank NO ONE. <!-- haha! so generous of you! JB: We should probably add some people here, no? -->
